%==================================
% (c) Jan Tulak, 2017
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction}\label{chap:introduction}
%----------------------------------------------------------------------


In projects with long life, even an initially clean code can become messy and complicated. Moreover when we speak about open-source projects where the original creators left yers ago and new people of various capabilities and knowledge continue the development.

In such a projects, new functionality is added to the existing code with minimal changes of the rest of the project. This may simplify the merging of these changes, as any responsible person can easily understand what the change does. But on the other side, in long term, it turns the code into so-called spaghetti.

The result is increasingly more difficult to maintain and test, and as a single functionality can be spread over many portions of the project, any change requires more and more attention and time.

XFSprogs, a package of tools for XFS filesystem, is such a project. While the filesystem itself is getting some strict eye from the Linux kernel community nowdays, the tools like {\tt mkfs.xfs}\footnote{Formats a partition as XFS.}, {\tt fsck.xfs}\footnote{Checks and repairs errors in an existing XFS filesystem.} and others are not so publicly exposed and gets a lot less attention. From my experience with working on this project, it happens that only one or two persons other than the author of a patch read it, and miss some subtle side effect the change has. Sometimes, the large set of tests XFS maintains captures this bug, sometimes it does not and it is noticed much later.

On this point, it is important to highlight that despite the naming
convention, each {\tt mkfs} tool is completely independent project and, for
example, {\tt mkfs.xfs} and {\tt mkfs.ext4} does not share any code except
system libraries.

Some parts of this code are more than 20 years old (see chapter 1 for more detailed history of XFS) and in the need of intensive cleaning. The tests suite (project xfstests) maintains hundreds of more or less complex tests, but these are limited in what they can detect as they usually work in this way: make a filesystem, then test that, so many errors in {\tt mkfs.xfs} are difficult to capture or notice. XFS also uses an automatic static analysis from Coverity, which is useful, but the project has no good data on the reliability of this analysis.

With the approval of David Chinner, then the maintainer of XFS, I began the
refactoring of mkfs.xfs, which was most in due. The goal of this work is to
repay the technical debt accumulated over the years. That means not only
fixing some long-known issues and cleaning particularly complex parts of
the code, but also making structural changes to minimize the amount of code
that has to be added or changed during the regular development (adding and
removing features of the filesystem\ldots). Which should slow the build-up
of the technical debt in the future.

After implementing these changes, the code should be verify how effective
the currently used tests and analysis are. Even if some testing and
analysis methods can be used only on a part of the code, the results, when
compared with other tools, still provides an estimate about the soundess
and completness of every used method.

At the same time, this work can be also seen as a review of how well can
various analysis and verification methods do on a real and in production
used code.

The refactoring was done in two parts. One set of changes was merged into
upstream in June 2016 (xfsprogs 4.7\footnote{The releasing of xfsprogs is
		tightly coupled with releases of XFS kernel module, which
		is part of Linux. Thus, xfsprogs uses the same version
		number that the respective XFS kernel module and Linux has,
because it is not expected to be generally used with any other version.}),
the other set is, at the time of writing, still waiting for a review. The
versions of xfsprogs at different stages of this work were:
\begin{itemize}
\item Before the beginning of refactoring -- xfsprogs 4.6.
\item After merging the first part -- xfsprogs 4.7.
\item Before merging the second part -- xfsprogs 4.11 at the time of writing.
\item After applying the second part -- not yet merged, changes only in a local repository.
\end{itemize}

This work is structured as follows: First, information about XFS and
mkfs.xfs are provided in the first chapter. In the following, third chapter, we look
at the refactoring done and discuss the changes. After that, another
chapter is dedicated to explaining formal analysis and verification,
describing common techniques, and we points out notable tools. And finally
in the fifth chapter, we discuss possible approaches for the analysis and
verification specifically of mkfs.xfs.

